---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CovCorTest
## Statistical Tests for Covariance and Correlation Matrices and their Structures

<!-- badges: start -->
<!-- badges: end -->

A compilation of tests for hypotheses regarding covariance and correlation matrices for one or more groups. The hypothesis can be specified through a corresponding hypothesis matrix and a vector or by choosing one of the basic hypotheses, while for the structure test, only the latter works. Thereby Monte-Carlo and Bootstrap-techniques are used, and the respective method must be chosen, and the functions provide p-values and mostly also estimators of calculated covariance matrices of test statistics.

## Installation

The offical version of CovCorTest can be installed using the R Console:
```{r, eval = FALSE}
install.packages("CovCorTest")
```

You can install the development version of CovCorTest from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("sjedhoff/CovCorTest")
```

## Structure of the package

The package is structures in tests regarding the covariance matrix and the correlation matrix and their structures. For this, each of the matrices, covariance and correlation, has eac three different test functions.
The best approach is to start with the simple functions: `TestCovariance_simple` and `TestCorrelation_simple`. These function take the dataset, the group sizes (when testing for multiple groups) and the hypothesis, which should be tested. Since the hypothesis can be chosen using a character string like "equal", no further knowledge about the matrices used to test the hypotheses is needed.
The advanced methods are `TestCovariance_base` and `TestCorrelation_base`, where the hypothesis is selected using a hypothesis matrix and a corresponding vector. This can be used to test all forms of hypotheses, but in-depth knowledge is necessary. 

The structures of the covariance and correlation matrices can be tested using `TestCovariance_structure`and `TestCorrelation_structure` respectively. Instead of a hypothesis, a structure can be selected using a string, which will then be tested.

## Example

```{r example}
library(CovCorTest)
## basic example code
```


## Literature

- Sattler, P. & Dobler, D. (2024). Testing for patterns and structures in covariance and correlation matrices. <em>arXiv preprint</em> https://arxiv.org/abs/2310.11799
- Sattler, P., Bathke, A.C. & Pauly, M. (2020). Testing Hypothese about Covariance Matrixes in general MANOVA Designs. <em>Journal of Statistical Planning and Inference</em> 219, 134-146 https://doi.org/10.1016/j.jspi.2021.12.001
- Sattler, P. & Pauly, M. (2024). Testing hypotheses about correlation matrices in general MANOVA designs. <em>TEST</em> 33, 496–516 https://doi.org/10.1007/s11749-023-00906-6


