---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CovCorTest

## Statistical Tests for Covariance and Correlation Matrices and Their Structures

<!-- badges: start -->
[![R-CMD-check](https://github.com/sjedhoff/CovCorTest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sjedhoff/CovCorTest/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/sjedhoff/CovCorTest/graph/badge.svg)](https://app.codecov.io/gh/sjedhoff/CovCorTest)
[![CRAN status](https://www.r-pkg.org/badges/version/CovCorTest)](https://CRAN.R-project.org/package=CovCorTest)
<!-- badges: end -->

`CovCorTest` provides statistical tests for hypotheses about covariance
matrices, correlation matrices, and structured covariance or correlation
matrices in one-sample and multiple-group designs.

The package offers predefined hypotheses as well as custom linear hypotheses
specified through a hypothesis matrix `C` and null vector `Xi`. Depending on
the selected test, p-values are approximated using bootstrap (`"BT"`), Monte
Carlo (`"MC"`), or Taylor-based Monte Carlo (`"TAY"`) procedures.

## Installation

Install the released version from CRAN:

```{r installation-cran, eval = FALSE}
install.packages("CovCorTest")
```

Install the development version from GitHub:

```{r installation-github, eval = FALSE}
# install.packages("devtools")
devtools::install_github("sjedhoff/CovCorTest")
```

## Main functions

- `test_covariance()` tests predefined or custom hypotheses about covariance
  matrices.
- `test_correlation()` tests predefined or custom hypotheses about correlation
  matrices.
- `test_covariance_structure()` tests covariance structures such as diagonal,
  compound symmetry, Toeplitz, autoregressive, and banded structures.
- `test_correlation_structure()` tests correlation structures such as
  diagonal, heterogeneous compound symmetry, heterogeneous Toeplitz,
  heterogeneous autoregressive, and banded structures.
- `test_combined()` compares the variances and correlations of exactly two
  groups.
- `get_hypothesis()` constructs `C` and `Xi` for a linear covariance model.

Observations must be stored in rows and variables in columns. Multiple groups
can be supplied as a list of matrices or data frames. Alternatively, they can
be combined in one matrix and their sample sizes supplied through `nv`.

## Quick start

The following example uses two groups from the `EEGwide` data included with
`MANOVA.RM`:

```{r example-data, eval = FALSE}
library(CovCorTest)

data("EEGwide", package = "MANOVA.RM")
vars <- colnames(EEGwide)[1:6]

data <- list(
  AD = EEGwide[
    EEGwide$sex == "M" & EEGwide$diagnosis == "AD",
    vars
  ],
  MCI = EEGwide[
    EEGwide$sex == "M" & EEGwide$diagnosis == "MCI",
    vars
  ]
)

nv <- vapply(data, nrow, integer(1))
```

Test equality of the covariance matrices:

```{r covariance-test, eval = FALSE}
set.seed(31415)

cov_result <- test_covariance(
  X = data,
  nv = nv,
  hypothesis = "equal",
  method = "BT",
  repetitions = 1000
)

cov_result
cov_result$pvalue
```

Test equality of the correlation matrices:

```{r correlation-test, eval = FALSE}
set.seed(31415)

cor_result <- test_correlation(
  X = data,
  nv = nv,
  hypothesis = "equal-correlated",
  method = "BT",
  repetitions = 1000
)

cor_result
```

Perform the combined test:

```{r combined-test, eval = FALSE}
set.seed(31415)

combined_result <- test_combined(
  X = data,
  nv = nv,
  repetitions = 1000
)

combined_result
combined_result$pvalue_Total
```

Test whether the covariance matrix of one group is diagonal:

```{r covariance-structure, eval = FALSE}
set.seed(31415)

structure_result <- test_covariance_structure(
  X = data$AD,
  structure = "diagonal",
  method = "BT",
  repetitions = 1000
)

structure_result
```

For banded structures, specify the number of off-diagonals that may be
nonzero:

```{r banded-structure, eval = FALSE}
set.seed(31415)

banded_result <- test_correlation_structure(
  X = data$AD,
  structure = "banded-toeplitz",
  bandwidth = 2,
  method = "BT",
  repetitions = 1000
)

banded_result
```

Other predefined hypotheses and structures are documented on the individual
function help pages:

```{r function-help, eval = FALSE}
?test_covariance
?test_correlation
?test_covariance_structure
?test_correlation_structure
```

At least 500 resampling repetitions are recommended. Larger values generally
provide more precise p-values but require more computation time.

## Custom hypotheses

Advanced users can supply a custom hypothesis matrix `C` and null vector `Xi`
directly. `get_hypothesis()` can construct them for a linear covariance model:

```{r custom-hypothesis, eval = FALSE}
d <- ncol(data$AD)
p <- d * (d + 1) / 2
diagonal_positions <- cumsum(c(1, d:2))

variance_component <- numeric(p)
variance_component[diagonal_positions] <- 1

V <- cbind(
  variance_component,
  1 - variance_component
)

hypothesis <- get_hypothesis(
  v0 = rep(0, p),
  V = V
)

set.seed(31415)

custom_result <- test_covariance(
  X = data$AD,
  C = hypothesis$hypothesis_matrix,
  Xi = hypothesis$hypothesis_vector,
  method = "MC",
  repetitions = 1000
)

custom_result
```

By default, the tests use `AM = 1`, which replaces the original hypothesis
matrix with a lower-dimensional companion matrix without changing the
ANOVA-type test statistic. Set `AM = 0` to disable this transformation.

## Citation

If you use `CovCorTest` in a scientific publication, please cite:

Sattler, P. and Jedhoff, S. (2025). Testing Hypotheses Regarding Covariance
and Correlation Matrices with the R Package CovCorTest. *arXiv:2507.03406*.  
<https://doi.org/10.48550/arXiv.2507.03406>

The citation and BibTeX entry are also available in R:

```{r citation, eval = FALSE}
citation("CovCorTest")
toBibtex(citation("CovCorTest"))
```

Depending on the procedures used, please also cite the corresponding
methodological paper.

## Methodological references

- **Covariance matrix hypotheses (`test_covariance()`):**  
  Sattler, P., Bathke, A. C. & Pauly, M. (2022). Testing hypotheses about
  covariance matrices in general MANOVA designs. *Journal of Statistical
  Planning and Inference* 219, 134–146.  
  <https://doi.org/10.1016/j.jspi.2021.12.001>

- **Correlation matrix hypotheses and the combined test
  (`test_correlation()`, `test_combined()`):**  
  Sattler, P. & Pauly, M. (2024). Testing hypotheses about correlation
  matrices in general MANOVA designs. *TEST* 33, 496–516.  
  <https://doi.org/10.1007/s11749-023-00906-6>

- **Covariance and correlation structure tests:**  
  Sattler, P. & Dobler, D. (2026). Testing for patterns and structures in
  covariance and correlation matrices. *Journal of Multivariate Analysis*
  211, 105517.  
  <https://doi.org/10.1016/j.jmva.2025.105517>

- **Alternative hypothesis matrices (`AM = 1`):**  
  Sattler, P. & Rosenbaum, M. (2025). Choice of the hypothesis matrix for
  using the ANOVA-type-statistic. *Statistics & Probability Letters* 219,
  110356.  
  <https://doi.org/10.1016/j.spl.2025.110356>
